The polynomial representation of the type $A_{n - 1}$ rational Cherednik algebra in characteristic $p \mid n$

### 2015-05-14 formalwriteup.tex
Fix some formatting

Let $T(\h \oplus \h^*)$ be the tensor algebra of $\h \oplus \h^*$. We choose $\hbar,c \in k$. Then we define the {\it rational Cherednik algebra of type $A$} $\HH_{\hbar,c}(\h)$ as the quotient of $k[S_n] \ltimes T(\h \oplus \h^*)$ by the relations
\begin{align*}
[x,x']=0, \quad [y,y' = 0], \quad [y,x] = \hbar\langle y,x\rangle - \sum_{s \in \mathcal{S}} c\langle y,\alpha_s\rangle\langle \alpha_s^\vee,x\rangle s
\end{align*}
 for all $x,x' \in \h^*, y,y' \in \h$. We can give $\HH_{\hbar,c}(\h)$ a $\mathbb{Z}$-grading by setting $\deg x=1$ for $x \in \h^*$, $\deg y = -1$ for $y \in \h$, and $\deg g=0$ for $g \in k[G]$. We get the PBW-type decomposition $\HH_{\hbar,c}(\h)=\Sym(\h) \otimes_k k[S_n] \otimes_k \Sym(\h^*)$ (\cite{EM}, section 3.2).

@@

To understand the action of $\HH_{\hbar,c}(\h)$ on $A$, we can use the PBW decompositions. The action of $\Sym(\h^*)$ on $A$ is by left multiplication; $k[S_n]$ acts by the diagonal action, and $\Sym(\h)$ acts via {\it Dunkl operators}. For $y \in \h$, the Dunkl operator $D_y$ acts on $A$ by:
\begin{align*}
D_y(f) = \hbar \partial_y f  - \sum_{s \in \mathcal{S}} c \frac{ ( y, \alpha_s )}{\alpha_s} (1-s). f.
\end{align*}

We choose bases for $\h,\h^*$ in the following way: let $V$ be the vector space spanned by $y_0,\dots,y_{n-1}$ and let $\h$ be the subspace spanned by $y_i-y_j$ for $i \ne j$; $S_n$ acts by permuting indices. Then if $x_0,\dots,x_{n-1}$ is the dual basis for $V^*$, we see that $\h^*$ is the span of $x_0,\dots,x_{n-1}$ under the relation $x_0+\dots+x_{n-1}=0$; alternatively we can consider $\h^*$ as the span of $x_0,\dots,x_{n-2}$ with $x_{n-1}$ defined as $-x_0-\dots-x_{n-2}$.  If $D_i$ is the Dunkl operator corresponding to $y_i$, the Dunkl operators for the elements of $\h$ are spanned by $D_i-D_j$ with $i \ne j$.

@@

Since $x_{n-1}=-x_0-\dots-x_{n-2}$, we see that $$\frac{\partial g}{\partial x_i}=-\frac{zg}{1-x_iz}+\frac{zg}{1-x_{n-1}z}$$ for all $0 \le i < n-1$. Let $$F(z) =\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right).$$ Note that $F(z)$ is symmetric and $F_a(z)=\frac{F(z)}{1-x_az}$ for all $a$. Then we see that for all $0 \le i < n-1$ we have 
\begin{align*}
\frac{\partial F}{\partial x_i}%&=\frac{\partial}{\partial x_i}\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right)\\
&=\left(\sum_{k=1}^{p-1}k\binom{c}{k}(g-1)^{k-1}\right)\frac{\partial g}{\partial x_i}\\
&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=1}^{p-1}k\binom{c}{k}(g-1)^{k-1}\right)g\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=0}^{p-2}(k+1)\binom{c}{k+1}(g-1)^{k}\right)g\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}\right)(g-1+1)\\
&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}\right)+\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k+1}\right)\right)\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}\right)+\left(\sum_{k=1}^{p-1}c\binom{c-1}{k-1}(g-1)^{k}\right)\right)\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(-c\binom{c-1}{p-1}(g-1)^{p-1}+\sum_{k=0}^{p-1}c\binom{c}{k}(g-1)^{k}\right)\\
&=\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)\left(-\binom{c-1}{p-1}(g-1)^{p-1}+F(z)\right).\\
\end{align*}
We define $$G_i(z)=\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)\left(-\binom{c-1}{p-1}(g-1)^{p-1}\right).$$ Then we see that $$\frac{\partial F}{\partial x_i}=G_i(z)+\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)F(z).$$

@@

Define $$a(z)=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z).$$ Recall that we showed $$F'(z)=V(z)+\left(\sum_j \frac{-cx_j}{1-x_jz}F(z)\right)=V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right)$$ where $[z^\ell]V(z)=0$ for $0 \le \ell \le p$. We then see that
\begin{align*}
a'(z)&=\frac{d}{dz}\left(\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z)\right)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F'(z)+\sum_{j=0}^{r-1}\frac{y_j}{(1-y_jz)\prod_{i=0}^{r-1}(1-y_iz)}F(z)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}\left(V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}a(z)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}-\left(\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)\right)\\\end{align*}
Since $[z^\ell]V(z)=0$ for $0 \le \ell \le p$ we see that $\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}=z^pb(z)$ for some power series $b(z)$. Therefore $$a'(z)=z^pb(z)-\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)$$ for some power series $b(z)$.

@@

For any integer $\ell$ we define the Laurent polynomial $$h_\ell(z)=z^{-\ell-r}\prod_i(1-y_iz).$$ Then
\begin{align*}
\frac{d}{dz}( h_\ell(z) a(z)) &= -(l + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_\ell(z) a(z) + h_\ell(z) a'(z) \\
&=  -(\ell + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_\ell(z) a(z) - h_\ell(z) \sum_j \frac{(m_j c - 1)y_i}{1 - y_ijz} a(z)  + z^ph_\ell(z) b(z)\\
&= -(\ell + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_im_i c}{1 - y_iz} h_\ell(z) a(z) + z^ph_\ell(z) b(z)\\
&= -\left((\ell + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_\ell(z) a(z) + z^ph_\ell(z) b(z).
\end{align*}

We note that $$-\left((\ell + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_\ell(z) $$ is a Laurent polynomial with lowest degree term $$- z^{-l-r-1} \left(\ell + r + \sum_i y_i m_i c\right) = - z^{-\ell-r-1}(\ell+ r)$$ and highest degree term 
$$- z^{-\ell - 1} (-1)^r  \prod_i y_i\left((\ell + r) - \sum_i m_i c\right).$$

@@

We can assume without loss of generality that $y_{r-1}=0$. Then if $r>1$, we can go through the same argument again with $y_0,\dots,y_{r-2}$. We are able to do this because our calculation of $F'(z)$ and $V(z)$ so that $z^p \mid V(z)$ relies only on the fact that $\sum m_iy_i=0$, which will still be true if we leave out $y_{r-1}=0$. We will then find that one of the $y_i$ with $i < r-1$ is $0$, which gives a contradiction since the $y_i$ are distinct. Therefore we must have $r=1$ and $y_0=0$.

@@

By Proposition 3.4 in \cite{BC1}, we see that the Hilbert series of $A/J$ is $\left(\frac{1-t^p}{1-t}\right)^{n-1}h(t^p)$ for some polynomial $h$ with nonnegative integer coefficients; since $I \subseteq J$, we see that $h_{A/I}(t) \ge h_{A/J}(t)$ coefficient-wise; however by this restriction of the form of $h_{A/J}(t)$, we see that the only possible choice for $h$ is $h(t)=1$. Therefore $h_{A/I}(t)=h_{A/J}(t)$, so $I=J$ and these $n-1$ generators generate the whole ideal $J$.

>>>>

Let $T(\h \oplus \h^*)$ be the tensor algebra of $\h \oplus \h^*$. We choose $\hbar,c \in k$. Then we define the {\it rational Cherednik algebra of type $A$} $\HH_{\hbar,c}(\h)$ as the quotient of $k[S_n] \ltimes T(\h \oplus \h^*)$ by the relations
\begin{align*}
[x,x']=0, \quad [y,y' = 0], \quad [y,x] = \hbar\langle y,x\rangle - \sum_{s \in \mathcal{S}} c\langle y,\alpha_s\rangle\langle \alpha_s^\vee,x\rangle s
\end{align*}
 for all $x,x' \in \h^*, y,y' \in \h$. We can give $\HH_{\hbar,c}(\h)$ a $\mathbb{Z}$-grading by setting $\deg x=1$ for $x \in \h^*$, $\deg y = -1$ for $y \in \h$, and $\deg g=0$ for $g \in k[G]$. We get the PBW-type decomposition $\HH_{\hbar,c}(\h)=\Sym(\h) \otimes_k k[S_n] \otimes_k \Sym(\h^*)$ (\cite[Section 3.2]{EM}).

@@

To understand the action of $\HH_{\hbar,c}(\h)$ on $A$, we can use the PBW decompositions. The action of $\Sym(\h^*)$ on $A$ is by left multiplication; $k[S_n]$ acts by the diagonal action, and $\Sym(\h)$ acts via {\it Dunkl operators}. For $y \in \h$, the Dunkl operator $D_y$ acts on $A$ by:
\begin{align*}
D_y(f) = \hbar \partial_y f  - \sum_{s \in \mathcal{S}} c \frac{ ( y, \alpha_s )}{\alpha_s} (1-s). f.
\end{align*}
We choose bases for $\h,\h^*$ in the following way: let $V$ be the vector space spanned by $y_0,\dots,y_{n-1}$ and let $\h$ be the subspace spanned by $y_i-y_j$ for $i \ne j$; $S_n$ acts by permuting indices. Then if $x_0,\dots,x_{n-1}$ is the dual basis for $V^*$, we see that $\h^*$ is the span of $x_0,\dots,x_{n-1}$ under the relation $x_0+\dots+x_{n-1}=0$; alternatively we can consider $\h^*$ as the span of $x_0,\dots,x_{n-2}$ with $x_{n-1}$ defined as $-x_0-\dots-x_{n-2}$.  If $D_i$ is the Dunkl operator corresponding to $y_i$, the Dunkl operators for the elements of $\h$ are spanned by $D_i-D_j$ with $i \ne j$.

@@

\begin{proof}
Since $x_{n-1}=-x_0-\dots-x_{n-2}$, we see that $$\frac{\partial g}{\partial x_i}=-\frac{zg}{1-x_iz}+\frac{zg}{1-x_{n-1}z}$$ for all $0 \le i < n-1$. Let $$F(z) =\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right).$$ Note that $F(z)$ is symmetric and $F_a(z)=\frac{F(z)}{1-x_az}$ for all $a$. Then we see that for all $0 \le i < n-1$ we have 
\begin{align*}
\frac{\partial F}{\partial x_i}%&=\frac{\partial}{\partial x_i}\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right)\\
&=\left(\sum_{k=1}^{p-1}k\binom{c}{k}(g-1)^{k-1}\right)\frac{\partial g}{\partial x_i}\\
&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=1}^{p-1}k\binom{c}{k}(g-1)^{k-1}\right)g\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=0}^{p-2}(k+1)\binom{c}{k+1}(g-1)^{k}\right)g\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}\right)(g-1+1)\\
&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}\right)+\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k+1}\right)\right)\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}\right)+\left(\sum_{k=1}^{p-1}c\binom{c-1}{k-1}(g-1)^{k}\right)\right)\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(-c\binom{c-1}{p-1}(g-1)^{p-1}+\sum_{k=0}^{p-1}c\binom{c}{k}(g-1)^{k}\right)\\
&=\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)\left(-\binom{c-1}{p-1}(g-1)^{p-1}+F(z)\right).\\
\end{align*}
We define $$G_i(z)=\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)\left(-\binom{c-1}{p-1}(g-1)^{p-1}\right).$$ Then we see that $$\frac{\partial F}{\partial x_i}=G_i(z)+\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)F(z).$$

@@

Define $$a(z)=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z).$$ Recall that we showed $$F'(z)=V(z)+\left(\sum_j \frac{-cx_j}{1-x_jz}F(z)\right)=V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right)$$ where $[z^\ell]V(z)=0$ for $0 \le \ell \le p$. We then see that
\begin{align*}
a'(z)&=\frac{d}{dz}\left(\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z)\right)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F'(z)+\sum_{j=0}^{r-1}\frac{y_j}{(1-y_jz)\prod_{i=0}^{r-1}(1-y_iz)}F(z)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}\left(V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}a(z)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}-\left(\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)\right).
\end{align*}
Since $[z^\ell]V(z)=0$ for $0 \le \ell \le p$ we see that $\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}=z^pb(z)$ for some power series $b(z)$. Therefore $$a'(z)=z^pb(z)-\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)$$ for some power series $b(z)$.

@@

For any integer $\ell$ we define the Laurent polynomial $$h_\ell(z)=z^{-\ell-r}\prod_i(1-y_iz).$$ Then
\begin{align*}
\frac{d}{dz}( h_\ell(z) a(z)) &= -(l + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_\ell(z) a(z) + h_\ell(z) a'(z) \\
&=  -(\ell + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_\ell(z) a(z) - h_\ell(z) \sum_j \frac{(m_j c - 1)y_i}{1 - y_ijz} a(z)  + z^ph_\ell(z) b(z)\\
&= -(\ell + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_im_i c}{1 - y_iz} h_\ell(z) a(z) + z^ph_\ell(z) b(z)\\
&= -\left((\ell + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_\ell(z) a(z) + z^ph_\ell(z) b(z).
\end{align*}
We note that $$-\left((\ell + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_\ell(z) $$ is a Laurent polynomial with lowest degree term $$- z^{-l-r-1} \left(\ell + r + \sum_i y_i m_i c\right) = - z^{-\ell-r-1}(\ell+ r)$$ and highest degree term 
$$- z^{-\ell - 1} (-1)^r  \prod_i y_i\left((\ell + r) - \sum_i m_i c\right).$$

@@

We can assume without loss of generality that $y_{r-1}=0$. Then if $r>1$, we can go through the same argument again with $y_0,\dots,y_{r-2}$. We are able to do this because our calculation of $F'(z)$ and $V(z)$ so that $z^p \mid V(z)$ relies only on the fact that $\sum m_iy_i=0$, which will still be true if we leave out $y_{r-1}=0$. We will then find that one of the $y_i$ with $i < r-1$ is $0$, which gives a contradiction since the $y_i$ are distinct. Therefore we must have $r=1$ and $y_0=0$. 
\end{proof}

@@

By Proposition 3.4 in \cite{BC1}, we see that the Hilbert series of $A/J$ is $\left(\frac{1-t^p}{1-t}\right)^{n-1}h(t^p)$ for some polynomial $h$ with nonnegative integer coefficients; since $I \subseteq J$, we see that $h_{A/I}(t) \ge h_{A/J}(t)$ coefficient-wise; however by this restriction of the form of $h_{A/J}(t)$, we see that the only possible choice for $h$ is $h(t)=1$. Therefore $h_{A/I}(t)=h_{A/J}(t)$, so $I=J$ and these $n-1$ generators generate the whole ideal $J$.
\end{proof}
